Evaluating Perceptual Judgements on 3D Printed Bar Charts

### 2024-01-22 jds.tex
Updates and Reviewer 1 corrections

---
title: "Evaluating Perceptual Judgements on 3D Printed Bar Charts"
author:
- first: Tyler
  first-init: T.
  last: Wiederich
  marker: 1
  footnote-id: 1
  email: twiederich2@huskers.unl.edu
- first: Susan
  first-init: S.
  last: VanderPlas
  marker: 1
  email: susan.vanderplas@unl.edu
short-title: Perceptual Judgements on 3D Printed Bar Charts
output:
  jds.rmd::pdf_article: 
    extra_dependencies: "subfig"
affiliation:
- institution: "University of Nebraska-Lincoln"
  prefix: Department of Statistics
  country: United States of America
  marker: 1
classoption:
- letterpaper
- inpress
- dvipsnames
preamble: |
  \usepackage{amsfonts,amsmath,amssymb,amsthm} 
  \usepackage{threeparttable}
  \usepackage{cleveref}
  \usepackage{booktabs} 
  \usepackage{lipsum} 
  \PassOptionsToPackage{dvipsnames}{xcolor} 
  \newcommand{\svp}[1]{{\textcolor{ForestGreen}{#1}}} 
  \newcommand{\tw}[1]{{\textcolor{Red}{#1}}}
abstract: "Graphical design principles typically recommend minimizing the dimensionality of a visualization - for instance, using only 2 dimensions for bar charts rather than providing a 3D rendering, because this extra complexity may result in a decrease in accuracy. This advice has been oft repeated, but the underlying experimental evidence is focused on fixed 2D projections of 3D charts. In this paper, we describe an experiment which attempts to establish whether the decrease in accuracy extends to 3D virtual renderings and 3D printed charts. We replicate the grouped bar chart comparisons in the  1984 Cleveland & McGill study, assessing the accuracy of numerical estimates using different types of 3D and 2D renderings."
keywords:
- graphics
- 3D bar charts
- 3D printing
bibliography: JDS2023.bib
footnote:
- content: Tyler Wiederich
  type: corresp
  id: 1
---

```{r, include = F}
knitr::opts_chunk$set(echo = F, dpi = 300, message =F, warning = F, error = F, dev = "ragg_png")

library(tidyverse)
library(RSQLite)
library(lme4)

con = dbConnect(SQLite(), '../2022-Tyler-3D_Graphical_Perception/experiment_interface/department.db')
dbListTables(con)
results = dbReadTable(con, 'results')
users = dbReadTable(con, 'user')
userMatrix = dbReadTable(con, 'userMatrix')
dbDisconnect(con)

valid.users <- users %>% 
  mutate(subject = paste0(nickname, participantUnique)) %>% 
  filter(age != 'Under 19')

#Fill in correct values for incorrect 3d graph kits
# 1 result, need to fill in fileID, graphtype, and plot
results = results %>% 
  mutate(fileID = ifelse(graphCorrecter %in% 'id-01/Type1-Rep01', 1, fileID),
         graphtype = ifelse(graphCorrecter %in% 'id-01/Type1-Rep01', 'Type1', graphtype),
         plot = ifelse(graphCorrecter %in% 'id-01/Type1-Rep01', '3dPrint', plot))

load('../2022-Tyler-3D_Graphical_Perception/experiment_interface/data/set85data.Rdata')
load('../2022-Tyler-3D_Graphical_Perception/experiment_interface/data/kits.Rdata')

trueRatios = datasets %>% 
  mutate(ratio.df = map(data, function(x)(x[!is.na(x[,'IDchr']),4])),
         trueRatio = map(ratio.df, function(x)(x[1,'Height'] / x[2,'Height']))) %>% 
  unnest(trueRatio) %>% 
  filter(fileID != 15) %>% 
  #mutate(Height = Height * 100) %>% 
  select(fileID, Height)

res_all = results %>% 
  left_join(trueRatios, by = 'fileID') %>% 
  mutate(response = log2(abs(byHowMuch - Height*100) + 1/8),
         subject = paste0(nickname, participantUnique),
         ratioLabel = round(100*Height, 1)) %>% 
  filter(subject %in% valid.users$subject) %>% 
  arrange(appStartTime)

res <- res_all %>% 
  filter(whichIsSmaller == 'Triangle (▲)',
         as.Date.POSIXct(appStartTime) <= '2023-05-22') #Only verified results from before the start of SDSS

```

# Introduction
Good communication requires both that the information be transmitted correctly and that the intended recipient be able to decode and understand the transmitted information accurately. In order to communicate effectively, we must use graphical forms that accurately convey information relevant to the task in question. In many cases, this means we must understand how accurately people can read quantitative information off of charts. While accuracy is not the only quantity of interest in graphical investigations [@hullmanPursuitErrorSurvey2019], it is an important factor in assessing the utility of many different data graphics.

The accuracy of graphical forms has been studied for almost a century [@vonhuhnFurtherStudiesGraphic1927;@eellsRelativeMeritsCircles1926;@croxtonGraphicComparisonsBars1932;@croxtonBarChartsCircle1927], as new ways of representing information evolve, we must revisit old studies to determine whether these representations have the same limitations as previous versions. This is particularly true in areas like graphics which are affected by the immense technological innovation in hardware and software which has taken place since the early 1990s.

## Elementary Graphical Tasks
@clevelandGraphical1984 established the comparative accuracy of different "elementary perceptual tasks" (EPTs).
Elementary Perceptual Tasks, according to these experiments, include assessing graphical elements such as position along a common scale, length, angle, and volume, and estimating the corresponding numerical value of these representations.
The study relied entirely on estimation accuracy, which may not always be relevant when extracting information from graphs. For example, estimation is less relevant when ordering values by size.
As a result of the Cleveland and McGill [-@clevelandGraphical1984] study, it is possible to assemble an ordering of perceptual accuracy for the elements of length, position, and angle. 
@heerCrowdsourcingGraphicalPerception2010b replicated some parts of @clevelandGraphical1984 in an online setting using Mechanical Turk, largely validating the results of the original study while demonstrating the utility of the Mechanical Turk platform for graphical testing.

The first experiment in @clevelandGraphical1984 (the position-length experiment), used five types of bar charts: two types of grouped bar charts and three types of stacked bar charts.
Each chart had two bars marked for comparison; participants were asked to determine which bar was smaller and give the  perceived ratio of the smaller bar to the larger bar. \Cref{fig:bar-types} shows the two types of grouped bar charts. 
We are primarily interested in the grouped bar charts (in part because 3D printing is not yet inexpensive enough to make moderate-scale stacked bar chart experiments viable), which consisted of two comparison bars which were either adjacent or in separate groups.
These grouped bar charts will be referenced as adjacent and separated graph types in this paper, respectively.

```{r bar-types}
#| fig-width: 4
#| fig-height: 4
#| fig-cap: "\\citet{clevelandGraphical1984} used two different types of grouped bar charts: comparisons between adjacent bars, and comparisons between separated bars. It is widely acknowledged that comparisons between separated bars (Type 3 comparisons, in Cleveland \\& McGill's terminology) are more difficult and error-prone."
#| out-width: ".49\\linewidth"
#| fig-show: "hold"
#| layout-ncol: 2
library(ggplot2)
library(ragg)
datasets$data[[1]] %>%
  ggplot(aes(x = Group, y = Height, group = GroupOrder)) + 
  geom_col(position = "dodge", fill = "white", color = "black") + 
  geom_point(aes(x = Group, y = 5, group = GroupOrder, shape = IDchr), size = 3, position = position_dodge(0.9)) + 
  guides(shape = "none") + 
  theme_bw() + 
  theme(axis.title = element_blank(), axis.text.x = element_blank()) + 
  ggtitle("Adjacent (Type 1)")

datasets$data[[1]] %>%
  ggplot(aes(x = factor(Order %%2), y = Height, group = GroupOrder)) + 
  geom_col(position = "dodge", fill = "white", color = "black") + 
  geom_point(aes(x = factor(Order %%2), y = 5, group = GroupOrder, shape = IDchr), size = 3, position = position_dodge(0.9)) +
  guides(shape = "none") + 
  theme_bw() + 
  theme(axis.title = element_blank(), axis.text.x = element_blank()) + 
  ggtitle("Separated (Type 3)")

```

## 3D Graphical Perception
Chart perception is often affected by the visual system's implicit assumption that visual stimuli are three-dimensional; after all, most of the visual input we process does come from a three-dimensional world, but charts are artificial and largely exist in two dimensions.
This occasionally causes problems: the line-width illusion[@vanderplasSignsSineIllusion2015;@daySineIllusion1991;@hofmannCommonAnglePlots2013], for instance, has been attributed to implicit 3D perception of 2D stimuli and can affect perception of error bands, candlestick plots, and Sankey diagrams (a simplified example is shown in @fig:sine-illusion). 
While this illusion is problematic in 2D, when the depth cues are actually present in a three-dimensional situation, the same perceptual heuristics contribute to accurate size perception. Thus, there is reason to think that perceptual accuracy may be dependent on the realism of the stimuli.

```{r sine-illusion}
#| message: false
#| error: false
#| warning: false
#| cache: false
#| fig-width: 4
#| fig-height: 4
#| out-width: "33%"
#| fig-cap: "An illustration of the sine illusion\\citep{vanderplasSignsSineIllusion2015}, also known as the line-width illusion. All vertical lines are the same length, but the lines in the middle of the curve appear to be much shorter. The illusion results when implicit perceptual corrections useful for perceiving the size of objects with depth are applied to 2D stimuli with no actual depth. As 3D heuristics occasionally cause inaccurate communication when applied to 2D objects, it is reasonable to think that there might be some situations where charts making use of a realistic third dimension might be perceived more accurately than their 2D equivalents."
#| fig-subcap:
#|    - "Sine Illusion"
#|    - "3D surface with similar features"
#|    - "Perceived vs. actual length"
#' Function to data frame
#'  
#' @param n number of values 
#' @param ell extent of vertical range
#' @param x range of horizontal values
#' @param f function
#' @param fp first derivative of function f
#' @param f2p second derivative of function f
#' @return data frame with n function values and derivatives along the x axis for a range given by x
#' @example
#' f <- function(x) 2*sin(x)
#' fp <- function(x) 2*cos(x)
#' f2p <- function(x) -2*sin(x)
#' dframe <- function.frame(n=50, ell=1, x=c(0,2*pi), f, fp, f2p)
#' require(ggplot2)
#' qplot(x, y=ystart, yend=yend, xend=x, geom="segment", data=dframe) +
#'    geom_point(aes(x,y))
function.frame <- function(n=200, ell=1, x, f, fprime, f2prime) {
  x <- seq(x[1], x[2], length=n)
  if (length(ell) != length(x)) ell <- rep(ell, length(x))
  y <- f(x)
  ystart <- y - ell/2
  yend <- y + ell/2
  # now correct for line illusion in vertical direction
  dy <- diff(range(y))
  dyl <- diff(range(c(ystart, yend)))
  # fprime and f2prime are sensitive to the aspect ratio of a plot
  # we represent it in framework of dy and dy+len
  # needs to be fixed by factor a:  
  a <- dy/dyl

  fp <- a*fprime(x)
  f2p <- a*f2prime(x)
  data.frame(x,y,ystart, yend, fp, f2p, a)
}

#' This function has to be split in parts and renamed
#' 
#' Right now this function is a convoluted mess.  We need to have separate
#' functions that 
#' (1) create a data frame for a given function,
#' its first and second derivatives and range in x
#' (this is what the function function.frame is supposed to do), 
#' (2) a trig transform function
#' (3) a quadratic approach function - better even, make (2) and (3) one function and use a parameter to decide on the method to use.
#' @example
#' f <- function(x) 2*sin(x)
#' fp <- function(x) 2*cos(x)
#' f2p <- function(x) -2*sin(x)
#' dframe <- createSine(n=50, ell=1, x=c(0,2*pi), f, fp, f2p)
#' require(ggplot2)
#' qplot(x=x, xend=x, y=y+ellx4.u, yend=y-ellx4.l, geom="segment", data=dframe, linetype=I(1)) + 
#'    theme_bw() + coord_fixed(ratio=1) + 
#'    xlab("x") + ylab("y") + 
#'    scale_x_continuous(breaks=seq(0, 2*pi, by=pi/2), minor_breaks=minor.axis.correction,
#'      labels=c("0", expression(paste(pi,"/2")), expression(pi), expression(paste("3",pi, "/2")), 
#'      expression(paste("2",pi))))

createSine <- function(n=200, len=1, f=f, fprime=fprime, f2prime=f2prime, a=0, b=2*pi) {
  #  if(getquadapprox & !is.function(f2prime)) f2prime <- function(x) -1*f(x) # for backwards compatibility
  x <- seq(a, b, length=n+2)[(2:(n+1))]
  ell <- rep(len, length=length(x))
  fx <- f(x)
  ystart <- fx - .5*ell
  yend <- fx + .5*ell

  # now correct for line illusion in vertical direction
  dy <- diff(range(fx))
  dx <- diff(range(x))
  # fprime works in framework of dx and dy, but we represent it in framework of dx and dy+len
  # needs to be fixed by factor a:  
#  a <- dy/(dy + len) 
  dyl <- diff(range(c(ystart, yend)))
  a <- dy/(dyl) 
  # ellx is based on the "trig" correction
  ellx <- ell / cos(atan(abs(a*fprime(x))))
  # ellx2 is based on linear approximation of f  
  ellx2 <- ell * sqrt(1 + a^2*fprime(x)^2)

  # make this a data frame - ggplot2 doesn't do well with floating vectors
  dframe <- data.frame(x=x, xstart=x, xend=x, y=fx, ystart=ystart, yend=yend, ell=ell, ellx = ellx, ellx2=ellx2)

  # third adjustment is based on quadratic approximation of f.
  # this needs two parts: correction above and below f(x)  
  #   if(getquadapprox & is.function(f2prime)){
  #     secseg <- do.call("rbind", lapply(dframe$x, function(i) getSecantSegment(i, dframe, f, fprime, f2prime)))
  #     dframe$ellx3.u <- secseg$sec.ell1
  #     dframe$ellx3.l <- secseg$sec.ell2
  #   }

  fp <- a*fprime(x)
  f2p <- a*f2prime(x)
  v <- 1 + fp^2
  lambdapinv <- 0.5*(sqrt(v^2-f2p*fp^2*ell) + v)    
#  lambdaminv <- -0.5*(sqrt(v^2+f2p*fp^2*ell) + v)
  lambdaminv <- 0.5*(sqrt(v^2+f2p*fp^2*ell) + v)

  dframe$ellx4.l <- 0.5*abs(lambdapinv)/sqrt(v)
  dframe$ellx4.u <- 0.5*abs(lambdaminv)/sqrt(v)
#  dframe$lambdam <- 1/lambdaminv
#  dframe$lambdap <- 1/lambdapinv
  fp <- fprime(x)
  f2p <- f2prime(x)
  v <- 1 + fp^2
#  lambda1 <- (-v + sqrt(v^2 - fp^2*f2p*ell))/(fp^2*f2p) # these two work
#  lambda2 <- (-v + sqrt(v^2 + fp^2*f2p*ell))/(fp^2*f2p) # these two work

  lambdaa <- -2*(sqrt(v^2 - f2p*fp^2*ell) + v)^-1 # identical to top
  lambdab <- 2*(sqrt(v^2 + f2p*fp^2*ell) + v)^-1

  dframe$lambdam <- lambdaa
  dframe$lambdap <- lambdab

  # qplot(a*(sqrt(1 + a^2*fprime(x)^2)*abs(lambdab))^-1, ellx4.u, data=dframe)
  # qplot(a*(sqrt(1 + a^2*fprime(x)^2)*abs(lambdaa))^-1, ellx4.l, data=dframe)
  # ggplot(aes(x, y+ell*(a*sqrt(v)*abs(lambdaa))^-1, xend=x, yend=y-ell*(a*sqrt(v)*abs(lambdab))^-1), data=dframe) + geom_segment()

  dframe
}

#' what is the function doing? - in one sentence
#' 
#' function description in a paragraph
#' @param x0 is a vector of locations for which the secant segments are supposed to be calculated
#' @param df df is a data frame
#' @param f function
#' @param fp first derivative of function f
#' @param f2p second derivative of function f
#' @return data frame consisting of .... what is the output of the function?
#' @example
#' # need an example here of how to use the function
getSecantSegment <- function(x0, df, f, fprime, f2prime){
# find the closest values of the grid df corresponding to x0
  # doesn't work right now
  ell.x0 <- sapply(x0, function(x) df$ell[which.min(x-df$x)])

  dy <- diff(range(with(df, c(y+ell.x0, y-ell.x0))))
  dx <- diff(range(df$x))
  a <- dx/dy

  #---- Actual Roots
  #   secSlope   <- -a/fprime(x0)
  #   temp       <- seq(min(df$x)-pi, max(df$x)+pi, .0001)
  #   leftend    <- temp[which.min(abs(f(temp) + ell.x1 - secSlope*(temp-x1)))]
  #   rightend   <- temp[which.min(abs(f(temp) - ell.x1 - secSlope*(temp-x1)))]
  #----

  #---- Approximation
  lambda1 <- (-(fprime(x0)^2 + 1) + sqrt((fprime(x0)^2 + 1)^2 - 2*fprime(x0)^2*f2prime(x0)*ell.x0))/(fprime(x0)^2*f2prime(x0))
  lambda2 <- (-(fprime(x0)^2 + 1) + sqrt((fprime(x0)^2 + 1)^2 + 2*fprime(x0)^2*f2prime(x0)*ell.x0))/(fprime(x0)^2*f2prime(x0))

  x1 <- lambda1*fprime(x0)+x0
  x2 <- lambda2*fprime(x0)+x0
  y1 <- f(x0)-lambda1
  y2 <- f(x0)-lambda2
  #----

  #---- Approximation V2 ----
  #   dx <- fprime(x0)
  #   dx1 <- 1+dx^2   
  #   ddx <- f2prime(x0)
  #   
  #   corr1 <- 1/a*(((dx1)-sqrt(dx1^2-2*ddx*(dx1-1)^2*ell.x0))/(ddx*dx))
  #   corr2 <- 1/a*(((dx1)-sqrt(dx1^2+2*ddx*(dx1-1)^2*ell.x0))/(ddx*dx))
  #   if(abs(dx)<.01) {
  #     dx1 <- 2
  #     corr1 <- corr2 <- 0
  #   }
  #   if(abs(dx)<.01){
  #     ddx <- 1
  #   }
  #   
  #   x1 = corr1 + x0
  #   x2 = corr2 + x0
  #   y1 = f(x1)+ell.x0
  #   y2 = f(x2)-ell.x0
  #----

  df2 <- data.frame(x=x0, y=f(x0), deriv=fprime(x0),
                    sec.xstart=x1, sec.xend = x2, 
                    sec.ystart=y1, sec.yend = y2,
                    ell.orig = 2*ell.x0)

  df2$sec.ell1 <- a*.5*ell.x0/with(df2, sqrt((sec.yend-y)^2+(sec.xend-x)^2))
  df2$sec.ell2 <- a*.5*ell.x0/with(df2, sqrt((y-sec.ystart)^2+(x-sec.xstart)^2))
  #   df2$sec.ell1 <- with(df2, sqrt((sec.yend-y)^2+(sec.xend-x)^2))
  #   df2$sec.ell2 <- with(df2, sqrt((y-sec.ystart)^2+(x-sec.xstart)^2))
  df2$type <- "Perceived Width"
  df2$a <- a
  return(df2)
}

correctx <- function(z, fprime, a=0, b=2*pi, w=1) {
  # w = 1/(shrink+1)
  const <- integrate(function(x) abs(fprime(x)), a, b)$value
  trans <- sapply(z, function(i) integrate(function(x) abs(fprime(x)), a, i)$value*(b-a)/const + a)
  # alternatively to the rowMeans, you could report back  
  # trans*(1-w) + z*w
  trans*w + z*(1-w)
}

f <- function(x) 2*sin(x)
fprime <- function(x) 2*cos(x)
f2prime <- function(x) -2*sin(x)

x <- seq(0, 2*pi, length=42)[2:41]
data <- do.call("rbind", lapply(seq(-.5, .5, 1), function(i) data.frame(x=x, y=3*sin(x), z=i)))

data.persp <- reshape2::acast(data, x~z, value.var="y")
x <- sort(unique(data$x))
y <- sort(unique(data$y))
z <- sort(unique(data$z))
dframe <- createSine(n = 50, len = 1, f=f, fprime=fprime, f2prime=f2prime)
dframe %>%
  ggplot(aes(x = xstart, xend = xend, y = ystart, yend = yend)) + geom_segment() + coord_fixed() + 
  theme_void()

par(mar = c(0, 0, 0, 0))
persp(x, z, data.persp,  xlab="", ylab="", zlab="", theta=0, phi=45, border="black", shade=.35, col="white", xlim=c(-pi/12, 2*pi+pi/12), ylim=c(-1, 1), scale=FALSE, box=FALSE, expand=2.5/pi, d=50) # , ltheta=0, lphi=-15

# persp(x, z, data.persp,  xlab="", ylab="", zlab="", theta=0, phi=45, border="black", shade=.35, col="white", xlim=c(-pi/12, 2*pi+pi/12), ylim=c(-1.75, 1.75), scale=FALSE, box=FALSE, expand=3/pi, d=3) # , ltheta=0, lphi=-15

dframe <- createSine(n = 150, len = 1, f=f, fprime=fprime, f2prime=f2prime)
dframe$ystartcts <- dframe$ystart
dframe$yendcts <- dframe$yend
dframe[1:150,c(2, 3, 5, 6)] <- NA
dframe[(1:15)*10-5, c(2, 3)] <- dframe[(1:15)*10-5, 1] 
dframe[(1:15)*10-5, 5] <- dframe[(1:15)*10-5, 4] - .5
dframe[(1:15)*10-5, 6] <- dframe[(1:15)*10-5, 4] + .5
dframe$type <- "Vertical Width"

idx <- which(!is.na(dframe$xstart))
dframe$ell <- dframe$ell/2
dframe.1 <- getSecantSegment(dframe$xstart[idx], dframe, f, fprime, f2prime)
dframe.1$x <- dframe$x[idx]
dframe.1$y <- dframe$y[idx]
dframe.1$ystartcts <- dframe$ystartcts[idx]
dframe.1$yendcts <- dframe$yendcts[idx]
names(dframe.1) <- c("x", "y", "deriv", "xstart", "xend", "ystart", "yend", "ell", "ell.quad1", "ell.quad2", "type", "a", "ystartcts", "yendcts")
dframe.1$vangle <- with(dframe.1, atan(deriv))
dframe <- bind_rows(dframe, dframe.1)
dframe$type <- factor(dframe$type)

ggplot(aes(x=x, y=y), data=dframe, colour=I("grey50")) + 
  geom_line(color = "grey") + theme_void() + 
  geom_line(aes(y=ystartcts), colour="grey50", alpha = .25) +
  geom_line(aes(y=yendcts), colour="grey50", alpha = .25) +
  geom_segment(data=subset(dframe, !is.na(type)), 
               aes(x=xstart, xend = xend, y=ystart, yend=yend, colour=type, linetype=type), linewidth=0.8)  + 
  xlab("") + ylab("")  +
  coord_equal(ratio=1) + scale_colour_manual("", values=c("blue", "grey30")) + 
  # geom_text(aes(label=paste("theta", "%~~%", round(abs(vangle)/pi*180), "^o", sep=""), 
                # x=pmax(xstart, xend)/2+x/2+.35 , y=y-sign(vangle)*.6+.02), colour="blue",
            # data=dframe.1, parse=TRUE, hjust=.9, vjust=.5, size=3) + 
  scale_x_continuous(breaks=seq(0, 2*pi, by=pi/2), 
                     labels=c("0", expression(paste(pi,"/2")), expression(pi), 
                              expression(paste("3",pi, "/2")), expression(paste("2",pi)))) +
  scale_linetype_manual("", values=c("11", "solid")) + 
  theme(legend.key.width = unit(3, "line"), plot.margin = unit(c(0,0,0,0), "cm"), legend.position=c(.25, .2))

```

The use of 3D graphics have been explored in multiple studies. 
@fisher_data_1997 explored user preference for 2D or 3D charts and found that subjects tended to prefer simpler 2D graphs when tasked with extracting information. 
@barfield_effects_1989 compared 2D and 3D graphs presented on paper and on computers, showing that the accuracy of subject answers depended on their skill level: novice subjects were more accurate with 2D paper graphs, while experienced managers were more accurate with 3D computer graphs. 
For both experience levels, participants were more confident in their answers when using 2D graphs. 
There are instances where 2D graphs perform better than 3D graphs, but there are times where 3D graphs may better encode information - for instance, when X and Y are used to represent spatial dimensions, it may be preferable to use a 3D chart to convey numerical information instead of using color, which is perceived much less accurately.
@brath_3d_2014 highlights the intrinsic attributes of 3D graphs and the benefits when used appropriately with other 3D elements such as lighting and correct portrayal of data attributes.

There is thus good reason to be wary of the use of three dimensions where only two are necessary to convey data. 
However, the situation is different now than it was in 1984 when Cleveland \& McGill published their seminal work; it has even changed since Heer's replication study [-@heerCrowdsourcingGraphicalPerception2010b] in 2010.  
Digital graphics have developed quickly, along with the hardware necessary to support these software developments. 
As a result, we have much more natural rendering of 3D objects virtually, and we can also print graphics in three dimensions, moving artificial charts into a more natural, physical setting.
As a result, it is reasonable to reconsider the use of 3D charts, not only because of new technological developments, but also because these charts provide the opportunity to make data graphics accessible to those with limited or absent vision [@fleetHowTactileGraphics2023].

In this paper, we discuss a study designed to examine Cleveland & McGill's experiments on grouped bar charts using modern graphics in two and three dimensions. This study lays the groundwork for additional empirical studies on the use of 3D graphics, rendered and 3D printed, for visualizing complex data. In order to explore perception of fully 3D graphics, it is prudent to start with the simplest possible 3D graphic: one in which the third dimension is not necessary, so that we can easily compare to two-dimensional representations without loss of information. <!-- This is included because of a graphics paper section comment that really annoyed me -->
In the next section we provide details about design and execution of our experiment, including the process of replicating stimuli from @clevelandGraphics1984 in order to create 2D, 3D projected, 3D rendered, and 3D printed bar graphs. The next section presents the results, and we conclude the paper by discussing this experiment in the context of existing work on the perception of 2D and 3D graphical elements.

# Methods

>>>>

\documentclass[letterpaper,inpress,dvipsnames]{jdsart}

\setcounter{page}{1}
\pubmonth{July}
\pubyear{2022}
\volume{xx}
\issue{xx}
\doi{0000}

\usepackage[utf8]{inputenc}
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\usepackage{subfig}

\usepackage{amsfonts,amsmath,amssymb,amsthm} 
\usepackage{threeparttable}
\usepackage{cleveref}
\usepackage{booktabs} 
\usepackage{lipsum} 
\PassOptionsToPackage{dvipsnames}{xcolor} 
\newcommand{\svp}[1]{{\textcolor{ForestGreen}{#1}}} 
\newcommand{\tw}[1]{{\textcolor{Red}{#1}}}

\begin{document}
\begin{frontmatter}

\title{Evaluating Perceptual Judgements on 3D Printed Bar Charts}
\runtitle{Perceptual Judgements on 3D Printed Bar Charts}

\author[1]{
  \inits{T.}
  \fnms{Tyler}
  \snm{Wiederich}  \thanksref{1}  \ead{twiederich2@huskers.unl.edu}}
\author[1]{
  \inits{S.}
  \fnms{Susan}
  \snm{VanderPlas}  \ead{susan.vanderplas@unl.edu}}

\thankstext[type=corresp,id=1]{Tyler Wiederich}
\address[1]{Department of Statistics, 
  \institution{University of Nebraska-Lincoln}, \cny{United States of America}}

\begin{abstract}
Graphical design principles typically recommend minimizing the dimensionality of a visualization - for instance, using only 2 dimensions for bar charts rather than providing a 3D rendering, because this extra complexity may result in a decrease in accuracy. This advice has been oft repeated, but the underlying experimental evidence is focused on fixed 2D projections of 3D charts. In this paper, we describe an experiment which attempts to establish whether the decrease in accuracy extends to 3D virtual renderings and 3D printed charts. We replicate the grouped bar chart comparisons in the 1984 Cleveland \& McGill study, assessing the accuracy of numerical estimates using different types of 3D and 2D renderings.
\end{abstract}

\begin{keywords}
\kwd{graphics}\kwd{3D bar charts}\kwd{3D printing}.
\end{keywords}

\end{frontmatter}

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

Good communication requires both that the information be transmitted correctly and that the intended recipient be able to decode and understand the transmitted information accurately. In order to communicate effectively, we must use graphical forms that accurately convey information relevant to the task in question. In many cases, this means we must understand how accurately people can read quantitative information off of charts. While accuracy is not the only quantity of interest in graphical investigations \citep{hullmanPursuitErrorSurvey2019}, it is an important factor in assessing the utility of many different data graphics.

The accuracy of graphical forms has been studied for almost a century \citep{vonhuhnFurtherStudiesGraphic1927, eellsRelativeMeritsCircles1926, croxtonGraphicComparisonsBars1932, croxtonBarChartsCircle1927}, as new ways of representing information evolve, we must revisit old studies to determine whether these representations have the same limitations as previous versions. This is particularly true in areas like graphics which are affected by the immense technological innovation in hardware and software which has taken place since the early 1990s.

\hypertarget{elementary-graphical-tasks}{%
\subsection{Elementary Graphical Tasks}\label{elementary-graphical-tasks}}

\citet{clevelandGraphical1984} established the comparative accuracy of different ``elementary perceptual tasks'' (EPTs).
Elementary Perceptual Tasks, according to these experiments, include assessing graphical elements such as position along a common scale, length, angle, and volume, and estimating the corresponding numerical value of these representations.
The study relied entirely on estimation accuracy, which may not always be relevant when extracting information from graphs. For example, estimation is less relevant when ordering values by size.
As a result of the Cleveland and McGill \citeyearpar{clevelandGraphical1984} study, it is possible to assemble an ordering of perceptual accuracy for the elements of length, position, and angle.
\citet{heerCrowdsourcingGraphicalPerception2010b} replicated some parts of \citet{clevelandGraphical1984} in an online setting using Mechanical Turk, largely validating the results of the original study while demonstrating the utility of the Mechanical Turk platform for graphical testing.

The first experiment in \citet{clevelandGraphical1984} (the position-length experiment), used five types of bar charts: two types of grouped bar charts and three types of stacked bar charts.
Each chart had two bars marked for comparison; participants were asked to determine which bar was smaller and give the perceived ratio of the smaller bar to the larger bar. \Cref{fig:bar-types} shows the two types of grouped bar charts.
We are primarily interested in the grouped bar charts (in part because 3D printing is not yet inexpensive enough to make moderate-scale stacked bar chart experiments viable), which consisted of two comparison bars which were either adjacent or in separate groups.
These grouped bar charts will be referenced as adjacent and separated graph types in this paper, respectively.

\begin{figure}
\includegraphics[width=.49\linewidth]{jds_files/figure-latex/bar-types-1} \includegraphics[width=.49\linewidth]{jds_files/figure-latex/bar-types-2} \caption{\citet{clevelandGraphical1984} used two different types of grouped bar charts: comparisons between adjacent bars, and comparisons between separated bars. It is widely acknowledged that comparisons between separated bars (Type 3 comparisons, in Cleveland \& McGill's terminology) are more difficult and error-prone.}\label{fig:bar-types}
\end{figure}

\hypertarget{d-graphical-perception}{%
\subsection{3D Graphical Perception}\label{d-graphical-perception}}

Chart perception is often affected by the visual system's implicit assumption that visual stimuli are three-dimensional; after all, most of the visual input we process does come from a three-dimensional world, but charts are artificial and largely exist in two dimensions.
This occasionally causes problems: the line-width illusion\citep{vanderplasSignsSineIllusion2015, daySineIllusion1991, hofmannCommonAnglePlots2013}, for instance, has been attributed to implicit 3D perception of 2D stimuli and can affect perception of error bands, candlestick plots, and Sankey diagrams (a simplified example is shown in \citet{fig:sine-illusion}).
While this illusion is problematic in 2D, when the depth cues are actually present in a three-dimensional situation, the same perceptual heuristics contribute to accurate size perception. Thus, there is reason to think that perceptual accuracy may be dependent on the realism of the stimuli.

\begin{figure}
\subfloat[Sine Illusion\label{fig:sine-illusion-1}]{\includegraphics[width=0.33\linewidth]{jds_files/figure-latex/sine-illusion-1} }\subfloat[3D surface with similar features\label{fig:sine-illusion-2}]{\includegraphics[width=0.33\linewidth]{jds_files/figure-latex/sine-illusion-2} }\subfloat[Perceived vs. actual length\label{fig:sine-illusion-3}]{\includegraphics[width=0.33\linewidth]{jds_files/figure-latex/sine-illusion-3} }\caption{An illustration of the sine illusion\citep{vanderplasSignsSineIllusion2015}, also known as the line-width illusion. All vertical lines are the same length, but the lines in the middle of the curve appear to be much shorter. The illusion results when implicit perceptual corrections useful for perceiving the size of objects with depth are applied to 2D stimuli with no actual depth. As 3D heuristics occasionally cause inaccurate communication when applied to 2D objects, it is reasonable to think that there might be some situations where charts making use of a realistic third dimension might be perceived more accurately than their 2D equivalents.}\label{fig:sine-illusion}
\end{figure}

The use of 3D graphics have been explored in multiple studies.
\citet{fisher_data_1997} explored user preference for 2D or 3D charts and found that subjects tended to prefer simpler 2D graphs when tasked with extracting information.
\citet{barfield_effects_1989} compared 2D and 3D graphs presented on paper and on computers, showing that the accuracy of subject answers depended on their skill level: novice subjects were more accurate with 2D paper graphs, while experienced managers were more accurate with 3D computer graphs.
For both experience levels, participants were more confident in their answers when using 2D graphs.
There are instances where 2D graphs perform better than 3D graphs, but there are times where 3D graphs may better encode information - for instance, when X and Y are used to represent spatial dimensions, it may be preferable to use a 3D chart to convey numerical information instead of using color, which is perceived much less accurately.
\citet{brath_3d_2014} highlights the intrinsic attributes of 3D graphs and the benefits when used appropriately with other 3D elements such as lighting and correct portrayal of data attributes.

There is thus good reason to be wary of the use of three dimensions where only two are necessary to convey data.
However, the situation is different now than it was in 1984 when Cleveland \& McGill published their seminal work; it has even changed since Heer's replication study \citeyearpar{heerCrowdsourcingGraphicalPerception2010b} in 2010.\\
Digital graphics have developed quickly, along with the hardware necessary to support these software developments.
As a result, we have much more natural rendering of 3D objects virtually, and we can also print graphics in three dimensions, moving artificial charts into a more natural, physical setting.
As a result, it is reasonable to reconsider the use of 3D charts, not only because of new technological developments, but also because these charts provide the opportunity to make data graphics accessible to those with limited or absent vision \citep{fleetHowTactileGraphics2023}.

In this paper, we discuss a study designed to examine Cleveland \& McGill's experiments on grouped bar charts using modern graphics in two and three dimensions. This study lays the groundwork for additional empirical studies on the use of 3D graphics, rendered and 3D printed, for visualizing complex data. In order to explore perception of fully 3D graphics, it is prudent to start with the simplest possible 3D graphic: one in which the third dimension is not necessary, so that we can easily compare to two-dimensional representations without loss of information.
In the next section we provide details about design and execution of our experiment, including the process of replicating stimuli from \citet{clevelandGraphics1984} in order to create 2D, 3D projected, 3D rendered, and 3D printed bar graphs. The next section presents the results, and we conclude the paper by discussing this experiment in the context of existing work on the perception of 2D and 3D graphical elements.

\hypertarget{methods}{%
\section{Methods}\label{methods}}

Our study is designed to replicate and expand upon the position-length experiment from Cleveland \& McGill as closely as possible, with the hope of being able to integrate this study with previously reported results in a consistent manner.
In this section, we discuss the replication process and the design of our version of this experiment.

\hypertarget{replicating-cleveland-and-mcgill}{%
\subsection{Replicating Cleveland and McGill}\label{replicating-cleveland-and-mcgill}}
